---
title: "Be the boss of your factors"
output:
  html_document:
    toc: true
    toc_depth: 4
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, collapse = TRUE, comment = "#>")
```

*Under development*

### Factors: where they fit in

We've spent alot of time working with big, beautiful data frame, like the Gapminder data. But we also need to manage the individual variables housed within.

Factors are the variable type that useRs love to hate. It is how we store truly categorical information in R. The values a factor can take on are called the **levels**. For example, the levels of the factor `continent` in Gapminder are are "Africa", "Americas", etc. and this is what's usually presented to your eyeballs by R. In general, the levels are friendly human-readable character strings, like "male/female" and "control/treated". But *never ever ever* forget that, under the hood, R is really storing integer codes 1, 2, 3, etc.

This [Janus](http://en.wikipedia.org/wiki/Janus)-like nature of factors means they are rich with booby traps for the unsuspecting but they are a necessary evil. I recommend you learn how to be the boss of your factors. The pros far outweigh the cons. Specifically in modelling and figure-making, factors are anticipated and accommodated by the functions and packages you will want to exploit.

The worst kind of factor is the stealth factor. The variable that you think of as character, but that is actually a factor (numeric!!). This is one of the classic gotchas in R. Check your variable types explicitly when things seem weird. Where do stealth factors come from? Base R has a burning desire to turn character information into factor. The most common place this happens is at data import via `read.table()` and friends. To shut this down, use `stringsAsFactors = FALSE` or -- even better -- use the tidyverse functions `read_csv()`, `read_tsv()`, etc.

Good articles about how the factor fiasco came to be:

  * [stringsAsFactors: An unauthorized biography](http://simplystatistics.org/2015/07/24/stringsasfactors-an-unauthorized-biography/) by Roger Peng
  * [stringsAsFactors = <sigh>](http://notstatschat.tumblr.com/post/124987394001/stringsasfactors-sigh) by Thomas Lumley

### The forcats package

This is a reboot of the [STAT 545 factor material from 2015 and before](block014_factors.html), in light of the very recent birth of the [forcats](https://blog.rstudio.org/2016/08/31/forcats-0-1-0/) package. 

forcats is a non-core package in the tidyverse. It is installed via `install.packages("tidyverse")`, but not loaded via `library(tidyverse)`. Load it yourself as needed via `library(forcats)`. Main functions start with `fct_`. There really is no coherent family of base functions that forcats replaces -- that's why it's such a welcome addition.

Currently this lesson will be mostly code vs prose. See the previous lesson for more discussion during the transition.

### Load tidyverse, forcats, gapminder

```{r}
library(tidyverse)
library(forcats)
library(gapminder)
```

### Factor inspection

Get to know your factor before you start touching it! It's polite.

```{r}
gapminder
str(gapminder)
str(gapminder$continent)
levels(gapminder$continent)
nlevels(gapminder$continent)
class(gapminder$continent)
summary(gapminder$continent)
```

Get a result similar to `dplyr::count()` but on a naked factor.

```{r}
gapminder %>% 
  count(continent)
fct_count(gapminder$continent)
```


### Dropping unused levels

`droplevels()` for operating on factors living in a data frame (or on a single factor).  
`fct_drop()` for operating on a factor directly.

```{r}
h_countries <- c("Egypt", "Haiti", "Romania", "Thailand", "Venezuela")
h_gap <- gapminder %>%
  filter(country %in% h_countries)
h_gap %>% str()
nlevels(h_gap$country)

## in data frame context
h_gap_dropped <- h_gap %>% 
  droplevels()
nlevels(h_gap_dropped$country)

## in a factor vector context
h_gap$country %>% levels()
h_gap$country %>% fct_drop() %>% levels()
```

### Change order of the levels, principled

Default order is alphabetical. Which is practically random, when you think about it! It is preferable to order the levels according to some principle:

  * Frequency. Make the most common level the first and so on.
  * Another variable. Order factor levels according to a summary statistic for another variable. Example: order Gapminder countries by life expectancy.

Order by frequency, forwards and backwards. Motivated by the downstream need to make tables and figures, esp. frequency barplots.

```{r}
## order by frequency
gapminder$continent %>%
  levels()
gapminder$continent %>% 
  fct_infreq() %>%
  levels() %>% head()
## backwards!
gapminder$continent %>% 
  fct_infreq() %>%
  fct_rev() %>% 
  levels() %>% head()
```

Order by another variable, forwards and backwards. This other variable is usually quantitative and you will order the factor accoding to a grouped summary. The factor is the grouping variable and the default summarizing function is `median()`.

```{r}
fct_reorder(gapminder$country, gapminder$lifeExp) %>% 
  levels() %>% head()
## order accoring to minimum life exp instead of median
fct_reorder(gapminder$country, gapminder$lifeExp, min) %>% 
  levels() %>% head()
## backwards!
fct_reorder(gapminder$country, gapminder$lifeExp, .desc = TRUE) %>% 
  levels() %>% head()
```

Example of why we reorder factor levels: often makes plots much better! When a factor is mapped to x or y, it should almost always be reordered by the quantitative variable you are mapping to the other one.

```{r include = FALSE, eval = FALSE}
boxplot(Sepal.Width ~ Species, data = iris)
boxplot(Sepal.Width ~ fct_reorder(Species, Sepal.Width), data = iris)
boxplot(Sepal.Width ~ fct_reorder(Species, Sepal.Width, .desc = TRUE), data = iris)
```

```{r alpha-order-silly, fig.show = 'hold', out.width = '49%'}
gap_asia_2007 <- gapminder %>% filter(year == 2007, continent == "Asia")
ggplot(gap_asia_2007, aes(x = lifeExp, y = country)) + geom_point()
ggplot(gap_asia_2007, aes(x = lifeExp, y = fct_reorder(country, lifeExp))) +
  geom_point()
```


Use `fct_reorder2()` when you have a line chart of a quantitative x against another quantitative y and your factor provides the color. This way the legend appears in some order as the data!

```{r legends-made-for-humans, fig.show = 'hold', out.width = '49%'}
h_countries <- c("Egypt", "Haiti", "Romania", "Thailand", "Venezuela")
h_gap <- gapminder %>%
  filter(country %in% h_countries) %>% 
  droplevels()
ggplot(h_gap, aes(x = year, y = lifeExp, color = country)) +
  geom_line()
ggplot(h_gap, aes(x = year, y = lifeExp,
                  color = fct_reorder2(country, year, lifeExp))) +
  geom_line() +
  labs(color = "country")
```

### Change order of the levels, "because I said so"

Sometimes you just want to hoist one or more level to the front. Because I said so. This resembles what we do when we move variables to the front with `dplyr::select(var1, var, everything())`.

```{r}
h_gap$country %>% levels()
h_gap$country %>% fct_relevel("Romania", "Haiti") %>% levels()
```

### Recode the levels

```{r}
i_gap <- gapminder %>% 
  filter(country %in% c("United States", "Sweden", "Australia")) %>% 
  droplevels()
i_gap$country %>% levels()
## oops United States is giving me trouble
i_gap$country %>%
  fct_recode("USA" = "United States", "Oz" = "Australia")%>% levels()

fct_count(gapminder$continent)
```

