---
title: "Basic care and feeding of data in R"
output:
  html_document:
    toc: true
    toc_depth: 4
  includes:
      before_body: include/clean-slate.html
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

### Buckle your seatbelt

*Ignore if you don't need this bit of support.*

Now is the time to make sure you are working in an appropriate directory on your computer, probably through the use of an [RStudio Project](block002_hello-r-workspace-wd-project.html). Enter `getwd()` in the Console to see current working directory or, in RStudio, this is displayed in the bar at the top of Console.

You should clean out your workspace. In RStudio, click on the "Clear" broom icon from the Environment tab or use *Session > Clear Workspace*. You can also enter `rm(list = ls())` in the Console to accomplish same.

Now restart R. This will ensure you don't have any packages loaded from previous calls to `library()`. In RStudio, use *Session > Restart R*. Otherwise, quit R with `q()` and re-launch it.

Why do we do this? So that the code you write is complete and re-runnable. If you return to a clean slate often, you will root out hidden dependencies where one snippet of code only works because it relies on objects created by code saved elsewhere or, much worse, never saved at all. Similarly, an aggressive clean slate approach will expose any usage of packages that have not been explicitly loaded. 

Finally, open a new R script and develop and run your code from there. In RStudio, use *File > New File > R Script*. Save this script with a name ending in `.r` or `.R`, containing no spaces or other funny stuff, and that evokes whatever it is we're doing today. Example: `cm004_data-care-feeding.r`.

### Get the Gapminder data

We will work with some of the data from the [Gapminder project](http://www.gapminder.org). I've released this as an R package, so we can install it from CRAN like so:

```{r eval = FALSE}
install.packages("gapminder")
```

Now load the package:

```{r}
library(gapminder)
```

### Meet your first data.frame

Whenever you have rectangular, spreadsheet-y data, your default data receptacle in R is a data.frame. Do not depart from this without good reason. data.frames are awesome because...

  * data.frames package related variables neatly together,
    - keeping them in sync vis-a-vis row order
    - applying any filtering of observations uniformly
  * most functions for inference, modelling, and graphing are happy to be passed a data.frame via a `data =` argument as the place to find the variables you're working on; the latest and greatest packages actually __require__ that your data be in a data.frame
  * data.frames -- unlike general arrays or, specifically, matrices in R -- can hold variables of different flavors (heuristic term defined later), such as character data (subject ID or name), quantitative data (white blood cell count), and categorical information (treated vs. untreated)

By loading the `gapminder` package, we now have access to a data.frame by the same name. Get an overview of this with `str()`, which displays the structure of an object. It will provide a sensible description of almost anything and, worst case, nothing bad can actually happen. When in doubt, just `str()` some of the recently created objects to get some ideas about what to do next.

```{r}
str(gapminder)
```

We could print the whole thing to screen (not so useful with datasets of any size) but it's nicer to look at the first bit or the last bit or a random snippet.

```{r include = FALSE}
set.seed(4561)
```

```{r}
head(gapminder)
tail(gapminder)
gapminder[sample(nrow(gapminder), 10), ]
```

More ways to query basic info on a data.frame. Note: with some of the commands below we're benefiting from the fact that even though data.frames are technically NOT matrices, it's usually fine to think of them that way and many functions have reasonable methods for both types of input.
```{r}
names(gapminder)
ncol(gapminder)
length(gapminder)
head(rownames(gapminder))
dim(gapminder)
nrow(gapminder)
#dimnames(gapminder) # ill-advised here ... too many rows
```

A statistical overview can be obtained with `summary()`

```{r}
summary(gapminder)
```

Although we haven't begun our formal coverage of visualization yet, it's so important for smell-testing dataset that we will make a few figures anyway. Here we use only base R graphics, which are very basic.

```{r first-plots-base-R}
plot(lifeExp ~ year, gapminder)
plot(lifeExp ~ gdpPercap, gapminder)
plot(lifeExp ~ log(gdpPercap), gapminder)
```

<!-- This is a non-sequitur here ... where came from originally?
> Sidebar on equals: A single equal sign `=` is most commonly used to specify values of arguments when calling functions in R, e.g. `group = continent`. It can be used for assignment but we advise against that, in favor of `<-`. A double equal sign `==` is a binary comparison operator, akin to less than `<` or greater than `>`, returning the logical value `TRUE` in the case of equality and `FALSE` otherwise. Although you may not yet understand exactly why, `subset = country == "Colombia"` restricts operation -- scatterplotting, in above examples -- to observations where the country is Colombia.
-->

Let's go back to the result of `str()` to talk about data.frames and vectors in R

```{r}
str(gapminder)
```

A data.frame is a special case of a *list*, which is used in R to hold just about anything. data.frames are the special case where the length of each list component is the same. data.frames are superior to matrices in R because they can hold vectors of different flavors (heuristic term explained below), e.g. numeric, character, and categorical data can be stored together. This comes up alot.

### Look at the variables inside a data.frame

To specify a single variable from a data.frame, use the dollar sign `$`. Let's explore the numeric variable for life expectancy.

```{r histogram-lifeExp}
head(gapminder$lifeExp)
summary(gapminder$lifeExp)
hist(gapminder$lifeExp)
```

The year variable is a numeric integer variable, but since there are so few unique values it also functions a bit like a categorical variable.

```{r}
summary(gapminder$year)
table(gapminder$year)
```

The variables for country and continent hold truly categorical information, which is stored as a *factor* in R.

```{r}
class(gapminder$continent)
summary(gapminder$continent)
levels(gapminder$continent)
nlevels(gapminder$continent)
```

The __levels__ of the factor `continent` are "Africa", "Americas", etc. and this is what's usually presented to your eyeballs by R. In general, the levels are friendly human-readable character strings, like "male/female" and "control/treated". But *never ever ever* forget that, under the hood, R is really storing integer codes 1, 2, 3, etc. Look at the result from `str(gapminder$continent)` if you are skeptical.

```{r}
str(gapminder$continent)
```

This [Janus](http://en.wikipedia.org/wiki/Janus)-like nature of factors means they are rich with booby traps for the unsuspecting but they are a necessary evil. I recommend you resolve to learn how to [properly care and feed for factors](block014_factors.html). The pros far outweigh the cons. Specifically in modelling and figure-making, factors are anticipated and accommodated by the functions and packages you will want to exploit.

Here we count how many observations are associated with each continent and, as usual, try to portray that info visually. This makes it much easier to quickly see that African countries are well represented in this dataset.
```{r tabulate-continent}
table(gapminder$continent)
barplot(table(gapminder$continent))
```

In the figures below, we see how factors can be put to work in figures. The `continent` factor is easily mapped into "facets" or colors and a legend by the `ggplot2` package. *Making figures with `ggplot2` is covered elsewhere so feel free to just sit back and enjoy these plots or blindly copy/paste.*

```{r factors-nice-for-plots, fig.show = 'hold', out.width = '49%'}
## install ggplot2 if you don't have it!
## install.packages("ggplot2")
library(ggplot2)
p <- ggplot(subset(gapminder, continent != "Oceania"),
            aes(x = gdpPercap, y = lifeExp)) # just initializes
p <- p + scale_x_log10() # log the x axis the right way
p + geom_point() # scatterplot
p + geom_point(aes(color = continent)) # map continent to color
p + geom_point(alpha = (1/3), size = 3) + geom_smooth(lwd = 3, se = FALSE)
p + geom_point(alpha = (1/3), size = 3) + facet_wrap(~ continent) +
  geom_smooth(lwd = 1.5, se = FALSE)
```

### `subset()` is a nice way to isolate bits of data.frames (and other things)

Logical little pieces of data.frames are useful for sanity checking, prototyping visualizations or computations for later scale-up, etc. Many functions are happy to restrict their operations to a subset of observations via a formal `subset =` argument. There is a stand-alone function, also confusingly called `subset()`, that can isolate pieces of an object for inspection or assignment. Although `subset()` can work on objects other than data.frames, we focus on that usage here.

The `subset()` function has a `subset =` argument (sorry, not my fault it's so confusing) for specifying which observations to keep. This expression will be evaluated within the specified data.frame, which is non-standard but convenient.

```{r}
subset(gapminder, subset = country == "Uruguay")
```

Contrast the above command with this one accomplishing the same thing:

```{r}
gapminder[1621:1632, ]
```

Yes, these both return the same result. But the second command is horrible for these reasons:

  * It contains [Magic Numbers](http://en.wikipedia.org/wiki/Magic_number_(programming)). The reason for keeping rows 1621 to 1632 will be non-obvious to someone else and that includes __you__ in a couple of weeks.
  * It is fragile. If the rows of `gapminder` are reordered or if some observations are eliminated, these rows may no longer correspond to the Uruguay data.

In contrast, the first command, using `subset()`, is self-documenting; one does not need to be an R expert to take a pretty good guess at what's happening. It's also more robust. It will still produce the correct result even if `gapminder` has undergone some reasonable set of transformations.

The `subset()` function can also be used to select certain variables via the `select` argument. It also offers unusual flexibility, so you can, for example, provide the names of variables you wish to keep without surrounding by quotes. I suppose this is mostly a good thing, but even the documentation stresses that the `subset()` function is intended for interactive use (which I interpret more broadly to mean data analysis, as opposed to programming).

You can use `subset =` and `select =` together to simultaneously filter rows and columns or variables.
```{r}
subset(gapminder, subset = country == "Mexico",
       select = c(country, year, lifeExp))
```

<!---
TO DO: CLEAN UP AND UN-COMMENT THESE EXERCISES FOR THE STUDENT
Let's get the data for just 2007.
How many rows?
How many observations per continent?
Scatterplot life expectancy against GDP per capita.
Variants of that: indicate continent by color, do for just one continent, do for multiple continents at once but in separate plots
```{r}
hDat <- subset(gapminder, subset = year == 2007)
str(hDat)
table(hDat$continent)
#xyplot(lifeExp ~ gdpPercap, hDat)
#xyplot(lifeExp ~ gdpPercap, hDat, group = continent, auto.key = TRUE)
#xyplot(lifeExp ~ gdpPercap | continent, hDat)
```
## if you want just some rows and/or just some variables, for inspection or to
## assign as a new object, use subset()
subset(gapminder, subset = country == "Cambodia")
subset(gapminder, subset = country %in% c("Japan", "Belgium"))
subset(gapminder, subset = year == 1952)
subset(gapminder, subset = country == "Uruguay", select = c(country, year, lifeExp))

plot(lifeExp ~ year, gapminder, subset = country == "Zimbabwe")
plot(lifeExp ~ log(gdpPercap), gapminder, subset = year == 2007)

## exercise:
## get data for which life expectancy is less than 32 years
## assign to an object
## how many rows? how many observations per continent?
--->

Many of the functions for inference, modelling, and graphics that permit you to specify a data.frame via `data = ` also offer a `subset =` argument that limits the computation to certain observations. Here's an example of subsetting the data to make a plot just for Colombia and a similar call to `lm` for fitting a linear model to just the data from Colombia.

```{r just-colombia}
p <- ggplot(subset(gapminder, country == "Colombia"),
            aes(x = year, y = lifeExp))
p + geom_point() + geom_smooth(lwd = 1, se = FALSE, method = "lm")
(minYear <- min(gapminder$year))
myFit <- lm(lifeExp ~ I(year - minYear), gapminder,
            subset = country == "Colombia")
summary(myFit)
```

### Review of data.frames and the best ways to exploit them

Use data.frames!!!

The most modern, slick way to work with data.frame is with the `dplyr` package. We will cover this extensively in the coming weeks. Here's material for 2014, though these will likely get an update:

  * [Introduction to dplyr](block009_dplyr-intro.html)
  * [`dplyr` functions for a single dataset](block010_dplyr-end-single-table.html)
  * [Cheatsheet for `dplyr` joins](bit001_dplyr-cheatsheet.html), i.e. "two table" operations
  
Work within your data.frames by passing them to the `data =` argument of functions that offer that. If you need to restrict operations, use the `subset =` argument. Do computations or make figures *in situ* -- don't create little copies and excerpts of your data. This will leave a cleaner workspace and cleaner code.

This workstyle leaves behind code that is also fairly self-documenting, e.g.,

```{r, eval = FALSE}
lm(lifeExp ~ year, gapminder, subset = country == "Colombia")
plot(lifeExp ~ year, gapminder, subset = country == "Colombia")
```

The availability and handling of `data =` and `subset =` arguments is broad enough-- though sadly not universal -- that sometimes you can even copy and paste these argument specifications, for example, from an exploratory plotting command into a model-fitting command. Consistent use of this convention also makes you faster at writing and reading such code. 

Two important practices

  * give variables short informative names (`lifeExp` versus "X5")
  * refer to variables by name, not by column number
  
This will produce code that is self-documenting and more robust. Variable names often propagate to downstream outputs like figures and numerical tables and therefore good names have a positive multiplier effect throughout an analysis.

If a function doesn't have a `data =` argument where you can provide a data.frame, you can fake it with `with()`. `with()` helps you avoid the creation of temporary, confusing little partial copies of your data. Use it -- possibly in combination with `subset()` -- to do specific computations without creating all the intermediate temporary objects you have no lasting interest in. `with()` is also useful if you are tempted to use `attach()` in order to save some typing. __Never ever use `attach()`. It is evil.__ If you've never heard of it, consider yourself lucky.

Example: How would you compute the correlation of life expectancy and GDP per capita for the country of Colombia? The `cor()` function sadly does not offer the usual `data =` and `subset =` arguments. Here's a nice way to combine `with()` and `subset()` to accomplish without unnecessary object creation and with fairly readable code.

```{r}
with(subset(gapminder, subset = country == "Colombia"),
     cor(lifeExp, gdpPercap))
```
