---
title: Exercises to test and solidy your data manipulation skills
output:
  html_document:
    toc: true
    toc_depth: 4
---

*NOTE: Not completed or used. It is a start on a set of data manipulation challenges, but I lost too much time tracking down a puzzle in the `spread()` example. It turned out to be a bug in `dplyr`. See [this issue](https://github.com/hadley/tidyr/issues/32) or [the one I opened and closed](https://github.com/hadley/tidyr/issues/42).*

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, collapse = TRUE)
```

```{r}
library(dplyr)
library(reshape2)
library(tidyr)
gdat <- read.delim("gapminderDataFiveYear.tsv")
```


### Aggregate or summarize

From this input:

```{r}
(hdat <- gdat %>%
   filter(country %in% c('France', 'Belgium', 'Nigeria', 'Japan'),
          year > 1996) %>%
   select(country, year, continent, lifeExp) %>%
   filter( (country == 'Japan') |
             (country == 'Belgium' & year == 2002) |
             (country == 'France' & year < 2005) |
             (country == 'Nigeria' & year > 2002)))
```

Make this output:

```{r include = FALSE}
idat <- hdat %>%
  group_by(country, continent) %>%
  summarize(nrows = n(),
            max_year = max(year),
            min_lifeExp = min(lifeExp))
```

```{r echo = FALSE}
knitr::kable(as.data.frame(idat))
```

### Cross-tabulate with holes

From `hdat` (code to produce given above)
```{r}
hdat
```

Make this output (it should be a data.frame):

```{r include = FALSE}
(jdat <- hdat %>%
   group_by(continent, year) %>%
   tally)

## tidyr::spread()
kdat <- jdat %>%
  ungroup %>% # necessary temporarily; fix coming to dplyr!
  spread(year, n)

## reshape2::dcast()
#dcast(jdat,continent ~ year, value.var = "n")
```

```{r echo = FALSE}
knitr::kable(as.data.frame(kdat))
```