---
title: 'cm008: Notes and Exercises'
date: '2017-09-28'
output: 
    html_document:
        toc: true
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
```

We're ahead of schedule! As a result, I'll talk about a variety of things first, then we'll do some exercises. 


## 1. Statistical Modelling in R

We'll look at typical data analyses using R.

**Note**: 

- You aren't expected to apply this is in your assignments! It's new to the course because we're ahead of schedule.
- It's OK if you've never heard of these statistical analyses. The point is that many model fitting procedures in R are similar.

Many statistical analyses in R follow a similar syntax.

### 1.1 Linear Regression

#### Model Fitting

You can run a linear regression in R with the `lm` function. Syntax:

```
lm(y ~ x1 + x2 + ... + xp, data=your_data_frame)
```

The first argument is a "formula" object in R. It's typically used in modelling to separate Y and X values. (In fact, you've seen this already in `ggplot`'s  `facet_wrap` and `facet_grid`)

Let's fit the regression curve that we see in this plot, using `lm`:

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    geom_point() +
    geom_smooth(method="lm") +
    scale_x_log10()
```

Here's the code:

```{r}
fit1 <- lm(lifeExp ~ log(gdpPercap), data=gapminder)
```

What does this `fit1` object look like?

```{r}
fit1
```

That's odd... what kind of R object is that??

```{r}
typeof(fit1)
```

It's a list, but R isn't presenting it that way. It just looks like a bunch of text, but it's not. Let's use the `lapply` function to uncover its true nature -- a list.

- NOTE: 
    - `lapply` loops over each component of a vector or list (_first argument_), applies a function to it (that you specify in the _second argument_), and outputs the function output in a list.
    - Let's do this for an `lm` fit to `head(gapminder)`, so that the output doesn't take up a lot of space. 

```{r}
fit1_small <- lm(lifeExp ~ log(gdpPercap), data=head(gapminder))
lapply(fit1_small, identity) 
```

Why isn't R printing out the list, then? Because it's a special type of list -- it's of class `"lm"`, something that the makers of the `lm` function decided. Whenever R encounters this object, it also has a special way of printing it to screen. 

This is the idea of the "object oriented" part of R -- something covered more in STAT 547 in the "R packages" section. 

#### Making predictions from the model

The `predict` function works on `"lm"` objects to make predictions. If you don't specify new data, it will make predictions using the existing X values. Let's look at the first six:

```{r}
predict(fit1) %>% head
```

How about plotted against the original X values? (Which was log gdpPercap)

```{r}
qplot(log(gapminder$gdpPercap), predict(fit1))
```

For fun... let's put this overtop of the scatterplot:

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    geom_point(alpha=0.1) +
    geom_point(y=predict(fit1), colour="red") +
    scale_x_log10()
```

You can predict with new data, too, as long as the data frame you enter has the same column names as your X values. 

```{r}
(my_newdata <- data.frame(gdpPercap=c(100, 547, 289)))
predict(fit1, newdata=my_newdata)
predict(fit1, newdata=filter(gapminder, country=="Canada"))
```


#### Extracting model characteristics

We can extract a bunch of things from the `lm` output.

- Regression coefficients? They're stored in the `$coefficients` part of the list. Or, use the `coeff` function.

```{r}
fit1$coefficients
coef(fit1)
```

- Residuals? They're stored in the `$residuals` part of the list. Or, use the `resid` function. (Let's only display the first six... but plot all of them!)

```{r}
fit1$residuals %>% head
resid(fit1) %>% head
qplot(log(gapminder$gdpPercap), resid(fit1)) +
    geom_hline(yintercept=0,
               linetype="dashed")
```

`lm` is kind of annoying in that not everything you might want is there. You can access more things using the `summary` function. What's printed to screen after `summary`, though, _is_ quite nice!

```{r}
(summ_fit1 <- summary(fit1))
```

You can see all sorts of things, like p-values, $R^2$ (and adjusted $R^2$ values), and standard errors. 

As before, this looks nice and all... but what the heck is this new object? Again, it's a list. Let's see its components (again, with the smaller fit, so that we don't take over all the space on the screen).

```{r}
summ_fit1_small <- summary(fit1_small)
typeof(summ_fit1_small)
lapply(summ_fit1_small, identity)  # Pry it open!!
```

There we have it. Now, what would you like to extract?:

- R-squared? R-squared adjusted? Okay:

```{r}
summ_fit1$r.squared
summ_fit1$adj.r.squared
```

- Estimated standard devaition of the random error term? Okay:

```{r}
summ_fit1$sigma
```

Where can we find the documentation for the components of _this_ list, though, if it's not in the documentation for `lm`? Look at the documentation of `summary.lm`. 

But wait! Why not just look at the documentation for `summary`? It's because `summary` is a generic function, and depends on the _class_ of object it's being applied to. If it's of class `"lm"`, then `summary.lm` is what's actually secretly run. Running `summary` on an object of class `"glm"`? R will secretly run `summary.glm` instead. 

PS: The `broom` package makes a lot of this easier and less cryptic. We won't go over it here. Check out [its vignette](https://cran.r-project.org/web/packages/broom/vignettes/broom.html).

### 1.2 Generalized Linear Models (like Logistic Regression)

We won't go over this in as much detail, because it's quite similar to `lm`. But if you want to run a Generalized Linear Model (GLM) -- such as logistic/binomial regression, or Poission regression -- just use the `glm` function.

Probably the biggest noteworthy difference is the `family` argument, specifying what type of regression you want to do. Syntax:

```
## Poisson regression:
glm(y ~ x1 + x2 + ... + xp, family=poisson, data=your_data_frame)
## Logistic (aka Binomial) regression:
glm(y ~ x1 + x2 + ... + xp, family=binomial, data=your_data_frame)
```

Its output looks similar to `lm`. It's also a list disguised as text. It also shows more when you use the `summary` function. It also works with the `predict` function. It also becomes tidier when used in conjunction with the `broom` package.

### 1.3 Others...

Here are some other packages/functions you might find useful to fit models:

- (Generalized) Mixed Effects Models
    - Two R packages are available: `lme4` and `nlme`. 
        - Check out [this](http://stats.stackexchange.com/questions/5344/how-to-choose-nlme-or-lme4-r-library-for-mixed-effects-models) discussion on Cross Validated for a comparison of the two packages.
    - I've found the function `glmer` in the `lme4` package to be fruitful.
- Kernel smoothing (i.e. fitting a "smoother"): check out the `loess` function.
- Generalized Additive Models: The `gam` function in _either_ the `gam` package or `mgcv` package.
- Robust linear regression: The `rlm` function in the `MASS` package is your friend.
- Regularized regression (GLM) (lasso, elastic net, or ridge regression): Use the `glmnet` function in the `glmnet` package.
    - PS: I _highly_ recommend this if you have more predictors/covariates/features than you know what to do with... this will weed out the unnecessary ones, _and_ produce a model with good prediction accuracy at the same time.


## 2. More `dplyr`

There's one more important thing with `dplyr` that you ought to know: applying `mutate` to a grouped tibble.

Remember applying `summarize` to a grouped tibble?

```{r}
gapminder %>%
    group_by(continent) %>% 
    summarize(mean_gdpPercap = mean(gdpPercap),
              n_countries    = length(gdpPercap))
```

Well, we can also apply `mutate` to each group, too. For example, let's calculate the growth in population since the first year on record _for each country_:

```{r}
gapminder %>% 
    group_by(country) %>% 
    mutate(pop_growth = pop - pop[1])
```

Notice that `dplyr` has retained the original grouping -- it hasn't pealed back one level of grouping. That's because there's still more than one row for each group!

How about growth compared to `1972`?

```{r}
gapminder %>% 
    group_by(country) %>% 
    mutate(pop_growth = pop - pop[year==1972])
```

In general, this type of "grouped mutation" is useful for _window functions_. What's that? Well, let's see other types of functions in R:

- __Vectorized Functions__: These take a vector, and operate on each component independently to return a vector of the same length. In other words, they work element-wise. 
    - Examples are `cos`, `sin`, `log`, `exp`, `round`. 
    - We don't need to `group_by` in order to `mutate` with these. 
- __Aggregate Functions__: These take a vector, and return a vector of length 1 -- as if "aggregating" the values in the vector into a single value.
    - Examples are `mean`, `sd`, `length`, `typeof`.
    - We use these in dplyr's `summarise` function.
- __Window Functions__: these take a vector, and return a vector of the same length _that depends on other values in the vector_. 
    - Examples are `lag`, `rank`, `cumsum`. 
    - See the [window-functions](https://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html) vignette for the `dplyr` package. 

## 3. More `ggplot`

(Have you seen the [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)? It contains __a lot__ of useful information on two pages!)

### 3.1 `theme` layers

You can change the look of a plot by adding a `theme` layer to your ggplot layers. This function does __not__ actually change the "nature" of the plot itself -- only the look! i.e., the so-called "non-data" type displays. 

Examples:
    - font
    - justification of titles
    - rotation of labels
    - background colour
    - line thickness
    - etc...

There are ["complete themes"](http://ggplot2.tidyverse.org/reference/ggtheme.html) that come with `ggplot2`, my favourite being `theme_bw` (I've grown tired of the default gray background, so `theme_bw` is refreshing).

Let's see an example:

```{r}
p1 <- ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    facet_wrap(~ continent) +
    geom_point(colour="#386CB0", alpha=0.2) +
    scale_x_log10()
p1 + theme_bw()
```

The general `theme` function gives you vast functionality... just check its documentation to see what things you can change. The arguments of `theme` follow a naming convention: `general.to.specific`

For example,
    - `axis.title` will allow you to change the font of the axis titles.
    - `axis.title.x` does the same, but focusses on the `x` axis.

__Note__: You _can't_ change the actual words this way! That's changing the nature of the plot, and not the _look_ of the plot. 

Once we've chosen an argument name, we need to specify its value. This is almost always the output of one of the following functions:

- `element_blank` (basically means replace with "nothing")
- `element_rect`: allows us to specify features of a _rectangle_.
- `element_line`: allows us to specify features of a _line_.
- `element_text`: allows us to specify font. 

Check out their documentation to see exactly how each feature is modified.

- __Example__: To `p1` above, do all of the following....
    - change the background strip colour to orange, 
    - change the axis titles' font sizes to 14, and 
    - change the panel titles' font sizes to 14 and bolded.

```{r}
p1 +
    theme(strip.background = element_rect(fill="orange"),
          axis.title = element_text(size=14),
          strip.text = element_text(size=14, face="bold"))
```

__Another example__: do the same, but in conjunction with the `theme_bw` (notice the order)

```{r}
## Correct:
p1 +
    theme_bw() +
    theme(strip.background = element_rect(fill="orange"),
          axis.title = element_text(size=14),
          strip.text = element_text(size=14, face="bold"))
## Incorrect:
p1 +
    theme(strip.background = element_rect(fill="orange"),
          axis.title = element_text(size=14),
          strip.text = element_text(size=14, face="bold")) +
    theme_bw()  # Overrides the previous `theme` call!
```


### 3.2 Modifying scales

Recall that we use some _scale_ to represent the _range of values that a variable takes_ in our data. `ggplot` chooses defaults for this scale, but we can change those.

Check out [this tutorial by Hadley Wickham](https://github.com/hadley/ggplot2-book/blob/master/scales.rmd) for scales. 

We can modify scales using a suite of functions that have the following naming convention: `scale_a_b`, where:

- `a` is the scale you want to change. Is it `colour`? `size`? `x` position?
- `b` typically speaks to the nature of the variable. `continuous` is your variable is continuous; `discrete` if discrete. But, could be other things in certain cases, like `log10`, or `date` if your variable consists of dates. `manual` is an option too.

__Examples__: `scale_x_continuous`, `scale_colour_discrete`, `scale_y_sqrt`.

As usual in ggplot, these functions are added as a layer. 

#### Useful arguments

There are many useful arguments here. Some are more self-explanatory than others. 

- `name`. The first argument. Indicate the name of the scale/legend here.
    - You can also use the `labs` function for X and Y axes, and even plot title.

```{r}
p1 + scale_y_continuous("Life Expectancy")
p1 + labs(x="GDP per capita", 
          y="Life Expectancy",
          title="My Plot")

ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    geom_point(aes(colour=continent),
               alpha=0.2) +
    scale_colour_discrete("Continents of\n the World")
```

- `breaks` (Typically of a continuous scale). Here, you get to specify _where_ along the scale you'd like to display a value. 
    - Numbers are on the scale of the data (such as population), not the geometric scale (such as a hex colour code, or number of pixels over in a plot). 

```{r}
## Log lines:
p1 + scale_x_log10(breaks=c((1:10)*1000,
                            (1:10)*10000))

p2 <- ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    geom_point(aes(colour=pop/10^9),
               alpha=0.2)
## Default breaks
p2 + scale_colour_continuous("Population\nin billions")
## New breaks
p2 + scale_colour_continuous("Population\nin billions",
                             breaks=seq(0,2,by=0.2))
```

- `labels`. Text to replace the data value labels. Most useful for discrete data.

```{r}
## Not a good idea:
p2 + scale_colour_continuous("My odd\npopulation\nscale",
                             breaks=c(0.2, 0.7, 1.2),
                             labels=c("small", "bigger", "big"))

## Discrete scale:
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    geom_point(aes(colour=continent),
               alpha=0.2) +
    scale_colour_discrete(labels=c("Af", "Am", "As", "Eu", "Oc"))
```

- `limits`. Lower and upper bounds of the data that you'd like displayed. Leave one as `NA` if you want to use the default.

```{r}
p1 + scale_y_continuous(limits=c(60,NA))
```

- `position`. Position of the scale. Also controllable using `theme` for the legend.

```{r}
p1 + scale_y_continuous(position="right")
p2 + theme(legend.position = "bottom")
```

## 4. Exercises. 

Practice these concepts in the following exercises.

__Exercise 1__: Suppose we want to calculate some quantity for each country in the `gapminder` data set. For each of the following quantities, indicate whether the function is _vectorized_, _aggregate_, or _window_, and use `dplyr` functions to calculate the specified variable.

- The change in population from 1962 to 1972.

```{r}
## It's an Aggregate function.
gapminder %>% 
    filter(year %in% c(1962, 1972)) %>% 
    arrange(year) %>% 
    group_by(country) %>% 
    summarize(pop_chg=diff(pop))
gapminder %>% 
    group_by(country) %>% 
    summarise(pop_chg=pop[year==1972]-pop[year==1962])
```


- The population, in billions.

```{r}
## It's a vectorized function.
gapminder %>%
    mutate(pop_in_bill = pop/10^9)
```


- The lagged gdpPercap
    - i.e., the value that appears for 1962 would be the gdpPercap in 1957 (the previous entry).
    - Hint: use the `lag` function, then filter out the `NA`'s created with the `is.na` function.

```{r}
## It's a window function.
gapminder %>% 
    group_by(country) %>% 
    arrange(year) %>% 
    mutate(lag_gdpPercap=lag(gdpPercap)) %>% 
    filter(!is.na(lag_gdpPercap))
```



__Exercise 2__: For the `gapminder` dataset, make a spaghetti plot showing the population trend (in millions) over time for each country, facetted by continent. Make as many of the following modifications as you can:

- Colour each line by the log maximum gdpPercap experienced by the country.
- Rotate the x-axis labels to be vertical.
- Remove the x-axis title.
- Give the legend an appropriate title.
- Put the y-axis on a log-scale. 
- Rename the y-axis title.
- Add more numbers along the y-axis.
- Give the plot a title, and center the title. 
- Only label the x axis with years 1950, 1975, and 2000.
- Move the colour scale to the bottom.
- Rename the colour legend

```{r}
gapminder %>% 
    group_by(country) %>% 
    mutate(max_gdpPercap=max(gdpPercap)) %>% 
    ggplot(aes(year, pop/10^6)) + 
    facet_wrap(~ continent) +
    geom_line(aes(group=country,
                  colour=log(max_gdpPercap)),
              alpha=0.25) +
    theme_bw() + # I added this because I like this theme.
    labs(title="Population Trends") +
    scale_y_log10("Population (millions)",
                  breaks=c(0.1, 1, 10, 100, 1000),
                  labels=c(0.1, 1, 10, 100, 1000)) +
    scale_x_continuous("", breaks=c(1950, 1975, 2000)) +
    scale_colour_continuous("log Maximum\nGDP per cap.") +
    theme(axis.text.x = element_text(angle=90),
          plot.title = element_text(hjust=0.5),
          legend.position = "bottom")
```
