---
title: "STAT 545 Class Meeting 10"
output:
    html_notebook:
        toc: true
        theme: cerulean
        number_sections: true
editor_options: 
  chunk_output_type: inline
---

# Learning Objectives


```{r setup, include=FALSE}
suppressPackageStartupMessages(library(tidyverse))
knitr::opts_chunk$set(fig.width=4, fig.height=3)
```

We are going to use a set of datasets that Joey put together for us. You can install them from github.

```{r}
# if you have not installed devtools, time to do it (uncomment following line)
# install.packages("devtools")
# uncomment and run the following line to get the `singer` package
devtools::install_github("JoeyBernhardt/singer")
```

## Join together

For the first part of the lecture, we are going to work a tad on some more joining operation.

The package `singer` comes with two smallish dataframe about songs and artists. Let's take a look at them

```{r}
library(singer)
data("songs")
songs
```

```{r}
data("locations")
locations
```

### Challenge 1

What flow would you use to get a dataframe with all the rows from `songs` where there is matching information the in `locations` dataframe?

```{r}

```

### Challenge 2

What flow would you use to get a dataframe with all the rows from `songs` where there is matching information the in `locations` dataframe, getting back a dataframe with `title`, `artist_name` and `year`?

```{r}

```

### Challenge 2

What command(s) would you use to get a dataframe with all the rows from `songs` where there is matching information the in `locations` dataframe, getting back a dataframe with `title`, `artist_name` and `year`?

```{r}

```

### Challenge 3

What flow would you use to get the number of releases (albums) present in this two small datasets per year?

```{r}

```


## Reshaping

Let's consider the bigger datagrame `artist_locations`

```{r}
data("singer_locations")
singer_locations
```


Let's suppose that we would like to do some plotting, showing the mean of `artist_hotttnesss`, `artist_familiarity` and `duration` for `year`. Let's select the appropriate part of the dataframe.

```{r}
hfd_y <- singer_locations %>%
  select(year, artist_hotttnesss, artist_familiarity, duration)
```


You may have heard that ggplot prefers long dataframes, what does that mean? Let's see with an example. The dataframe we have is **wide**: that means that the observation presents the values of interest on different columns.

```{r}
hfd_y
```

To have in on a long format, we can use `tidyr`'s function `gather()`. Let's see:

```{r}
hfd_y_long <- hfd_y %>%
  gather(key = "measure", value = "Value", artist_hotttnesss:duration)
```

### micro challenge

Observe the number of rows and the first entries: what do you think it happened?


---

We can go back to the wide format using `tidyr`'s function `spread()`. Yet, be braced for a problem:

```{r}
hfd_y_long %>%
  spread(measure, Value)
```

What did go wrong here? As we had more than one song per year, when we made it long we lost some information: we don't know how to reassemble the wide table. Let's try again, but with additional wizardry: remember that if you want to gather and then spread, you need to have in the dataframe some column(s) uniquely identifying each row. In this case `song_id` does the magic.


```{r}
hfd_y_unique <- singer_locations %>%
  select(song_id, year, artist_hotttnesss, artist_familiarity, duration)
```

Gather:

```{r}
hfd_y_unique_long <- hfd_y_unique %>%
  gather(key = "measure", value = "Value", artist_hotttnesss:duration)
```

and then spread

```{r}
hfd_y_unique_long %>%
  spread(measure, Value)
```