---
title: "Syllabus"
output:
  html_document:
    includes:
      after_body: include/nothing.html
---

```{r include = FALSE}
library(dplyr)
library(stringr)
library(readr)
library(tidyr)
```

```{r include = FALSE}
rawdat <-
  read_csv(file.path("course-admin", "class-meetings", "class-meetings.csv"))
cdat <- rawdat %>%
  mutate(ugly_date = ISOdate(2016, match(month, tolower(month.name)), day),
         date = sprintf("%s-%02d %s",
                        tolower(months(ugly_date, abbreviate = TRUE)),
                        day, dow),
         notes = ifelse(grepl("^cm", cm) & raw_notes != "",
                        sprintf("%s: %s", cm, raw_notes), raw_notes))
#cdat %>% View

## link to the class page of any class meetings where such exists
## AND I want to link
## year 1: we just filtered for presence of .html file
## year 2: yes merge those but also filter on link_it for gradual reveal
cmdat <- data_frame(html = list.files(pattern = '^cm.*html$'))
cmdat <- cmdat %>% mutate(cm = str_extract(html, "cm[0-9]+"))
cdat <- cdat %>% full_join(cmdat)

cdat <- cdat %>% replace_na(list(html = ""))
cdat$notes[cdat$link_it] <-
  str_replace(cdat$notes[cdat$link_it],
              "(cm[0-9]{3})", '<a href="LINK">\\1</a>')
cdat <- cdat %>% mutate(notes = str_replace(notes, "LINK", html))
dat <- cdat %>% split(cdat$course)
dat <- lapply(dat, 'rownames<-', NULL)
```

In early September 2016, this reflects what happened in 2015. As the semester unfolds, we update with 2016 plans and reality. Overview:

  * STAT 545 = 7 weeks
  * STAT 547 = 6 weeks
  * First 7 + 2 = 9 weeks is general data wrangling, exploration, analysis
  * Last 4 weeks = 4 one-week self-contained intermediate topics (order likely to change):
    - Build your first R package
    - Build a Shiny app
    - Get data from the web (scraping and APIs)
    - Automate tasks and pipelines

<!-- unholy hack to make following two tables less wide and the same wide -->
<style type="text/css">
table {
   max-width: 50%;
}
</style>

#### STAT 545A (2015 plan ... gradually overwritten by 2016)

```{r results = 'asis', echo = FALSE}
knitr::kable(dat[["STAT545A"]] %>% select(date, notes))
```

  * [Homework 01](hw01_edit-README.html) Edit `README.md`, use R Markdown to explore Gapminder
  * [Homework 02](hw02_explore-gapminder-dplyr.html) Explore Gapminder as a tibble and starting using dplyr and ggplot2
  * Homework 03 Manipulate, summarize, visualize Gapminder with the tidyverse and ggplot2
  * Homework 04 ~~Use your own functions with split-apply-combine and functional programming~~ *no, this will change*
  * Homework 05 Prove you are in control of factors and figures. Clean up your repo.
  * Homework 06 *did not exist; it's "clean up your repo" and was folded into homework 05*

#### STAT 547M (tentative = mostly what we did in 2015)

```{r results = 'asis', echo = FALSE}
knitr::kable(dat[["STAT547M"]] %>% select(date, notes))
```

  * Homework 06 *did not exist; it's "clean up your repo" and was folded into homework 05*
  * Homework 07 Clean and wrangle some wild-caught data, such as the `candy` survey or the `bellybutton` data
  * [Homework 08](hw08_package.html) Write an R package
  * [Homework 09](hw09_automation.html) Automate a data analysis pipeline
  * [Homework 10](hw10_data-from-web.html) Get data from the web
  * [Homework 11](hw11_shiny.html) Build a Shiny apps
