---
title: "Stat 545 getting data from the Web"
author: "Andrew MacDonald"
output: 
    html_document:
        toc: true
        toc_depth: 4
---

```{r message=FALSE}
library(dplyr)
library(knitr)
library(devtools)
```

# Introduction

**All this and more is described at the [rOpenSci repository of R tools for interacting with the internet]( https://github.com/ropensci/webservices)**

There are many ways to obtain data from the Internet; let's consider four categories:

* *click-and-download* on the internet as a "flat" file, such as .csv, .xls
* *install-and-play* an API for which someone has written a handy R package
* *API-query* published with an unwrapped API
* *Scraping* implicit in an html website

# Click-and-Download

In the simplest case, the data you need is already on the internet in a tabular format. There are a couple of strategies here:

* use `read.csv` or `readr::read_csv` to read the data straight into R.

* use the command line program `curl` to do that work, and place it in a Makefile or shell script (see the Make lesson for how to do that).

The second case is most useful when the data you want has been provided in a format that needs cleanup. For example, the World Value Survey makes several datasets available as Excel sheets. The safest option here is to download the `.xls` file, then read it into R with `readxl::read_excel()` or something similar. An exception to this is data provided as Google Spreadsheets, which can be read straight into R using the [`googlesheets`](https://github.com/jennybc/googlesheets) package.

### from ropensci web services page:

* `downloader::download()` for SSL
* `curl::curl()` for SSL.
* `httr::GET` data read this way needs to be parsed later with read.table
* `rio::import()` can "read a number of common data formats directly from an https:// URL".  Isn't that very similar to the previous?


What about packages that install data?

# Data supplied on the web

Many times, the data that you want is not already organized into one or a few tables that you can read directly into R. More frequently, you find this data is given in the form of an API. **A**pplication **P**rogramming **I**nterfaces (APIs) are descriptions of the kind of requests that can be made of a certain piece of software, and descriptions of the kind of answers that are returned. Many sources of data -- databases, websites, services -- have made all (or part) of their data available via APIs over the internet. Computer programs ("clients") can make requests of the server, and the server will respond by sending data (or an error message). This client can be many kinds of other programs or websites, including R running from your laptop.

# install-and-play 

Many common web services and APIs have been "wrapped", i.e. R functions have been written around them which send your query to the server and format the response.

Why do we want this?

* provenance
* reproducible
* updating
* ease
* scaling

## Sightings of birds: `rebird`

[Rebird](https://github.com/ropensci/rebird) is an R interface for the [ebird](http://ebird.org/content/ebird/) database. Ebird lets birders upload sightings of birds, and allows everyone access to those data.

```{r eval=FALSE}
install.packages("rebird")
```

```{r message=FALSE}
library(rebird)
```

### Search birds by geography

We can use the function `ebirdgeo` to get a list for an area. (Note that South and West are negative):
```{r results='asis'}
vanbirds <- ebirdgeo(lat = 49.2500, lng = -123.1000)
vanbirds %>%
  head %>%
	kable
```
**Note**: Check the defaults on this function. e.g. radius of circle, time of year.

We can also search by "region", which refers to short codes which serve as common shorthands for different political units. For example, France is represented by the letters **FR**
```{r eval=FALSE}
ebirdregion("FR")
```
(note that the link in the help file leads to a dead link (as I write this on 24 Nov) but you can probably use the codes from geonames, below)

Find out *WHEN* a bird has been seen in a certain place! Choosing a name from `vanbirds` above (the Bald Eagle):

```{r eval=FALSE}
ebirdgeo(species = 'Haliaeetus leucocephalus', lat = 42, lng = -76)
```

`rebird` **knows where you are**:
```{r eval=FALSE}
ebirdgeo(species = 'Buteo lagopus')
```


## Searching geographic info: `geonames`
```{r message=FALSE}
#install.packages("rjson")
#install_github("ropensci/geonames")

library(geonames)
```

There are two things we need to do to be able to use this package to access the geonames API

1. go to [the geonames site](www.geonames.org/login/) and register an account. 
2. click [here to enable the free web service](http://www.geonames.org/enablefreewebservice)
3. Tell R your geonames username:

```{r eval = FALSE}
options(geonamesUsername="?????")
```

What can we do? get access to lots of geographical information via the various "web services" see [here](http://www.geonames.org/export/ws-overview.html)

```{r}
countryInfo <- GNcountryInfo()
```

```{r results='asis'}
countryInfo %>%
	head %>%
	kable
```

This country info dataset is very helpful for accessing the rest of the data, because it gives us the standardized codes for country and language.  

### remixing `geonames` and `rebird`:
What are the cities of France?
```{r results='asis'}
francedata <- countryInfo %>%
	filter(countryName == "France")

frenchcities <-	with(francedata,
			 GNcities(north = north, east = east, south = south,
			 				 west = west, maxRows = 500))
```

How many birds have been seen in France? Use the `countryCode` from the geonames data to get bird data from rebird!

```{r results='asis'}
francebirds <- countryInfo %>%
	filter(countryName == "France")


allbirds <- ebirdregion(francebirds$countryCode)  ## or perhaps fipsCode?

nrow(allbirds)
```

### Wikipedia searching 

Geonames also helps us search Wikipedia.
```{r results='asis'}
GNwikipediaSearch("London") %>%
  select(-summary) %>%
	head %>%
	kable
```

We can use geonames to search for georeferenced Wikipedia articles. Here are those within 20 Km of Rio de Janerio, comparing results for English-language Wikipedia (`lang = "en"`) and Portuguese-language Wikipedia (`lang = "pt"`):
```{r}
rio_english <- GNfindNearbyWikipedia(lat = -22.9083, lng = -43.1964,
																		 radius = 20, lang = "en", maxRows = 500)
rio_portuguese <- GNfindNearbyWikipedia(lat = -22.9083, lng = -43.1964,
																				radius = 20, lang = "pt", maxRows = 500)

nrow(rio_english)
nrow(rio_portuguese)
```


## Searching the Public Library of Science: `rplos`
PLOS ONE is an open-access journal. They allow access to an impressive range of search tools, and allow you to obtain the full text of their articles. 

```{r eval=FALSE}
install.packages("rplos")
## Do this only once:
```

```{r}
library(rplos)
```
Immediately we get a message. It's a link to the [tutorial on the Ropensci website!](http://ropensci.org/tutorials/rplos_tutorial.html). How nice :)

* We also get instructions to [create a PLOS account](https://register.plos.org/ambra-registration/register.action)
* Then go to Article Level Metrics http://alm.plos.org/
* click on your name to find your key.

```{r eval = FALSE}
Sys.setenv(PlosApiKey = "Paste your Key in here!!")
key <-  Sys.getenv("PlosApiKey")
```


### alternate strategy for keeping keys: `.Rprofile`
**Remember to protect your key! it is important for your privacy. You know, like a key**
* Now we follow the ROpenSci [tutorial on API keys](https://github.com/ropensci/rOpenSci/wiki/Installation-and-use-of-API-keys)
* ___Add `.Rprofile` to your `.gitignore` !!___
* Make a `.Rprofile` file [windows tips](http://cran.r-project.org/bin/windows/rw-FAQ.html#What-are-HOME-and-working-directories_003f) [mac tips](http://cran.r-project.org/bin/macosx/RMacOSX-FAQ.html#The-R-Console)
* Write the following in it:

```{r eval = FALSE}
options(PlosApiKey = "YOUR_KEY")
```

* restart R (e.g. reopen your Rstudio project)

This code adds another element to the list of options, which you can see by calling `options()`. Part of the work done by `rplos::searchplos()` and friends is to go and obtain the value of this option with the function `getOption("PlosApiKey")`.  This indicates two things: 
  1. Spelling is important when you set the option in your `.Rprofile`
  2. you can do a similar process for an arbitrary package or key. For example:

```{r eval = FALSE}
## in .Rprofile
options("this_is_my_key" = XXXX)

## later, in the R script:
key <- getOption("this_is_my_key")
```

This is a simple means to keep your keys private, especially if you are sharing the same authentication across several projects.

### A few timely reminders about your .Rprofile

```r
print("This is Andrew's Rprofile and you can't have it!")
options(PlosApiKey = "XXXXXXXXX")
```

* Note that it must end with a blank line!
* It lives in the project's working directory, i.e. the location of your `.Rproj`
* It must be gitignored

Remember that using `.Rprofile` makes your code un-reproducible.  In this case, that is exactly what we want!  

## Searching PLOS
Let's do some searches:
```{r eval=FALSE}
searchplos(q= "Helianthus", fl= "id", limit = 5)
```

```{r eval=FALSE}
searchplos("materials_and_methods:France",
					 fl = "title, materials_and_methods", key = key)
lat <- searchplos("materials_and_methods:study site",
									fl = "title, materials_and_methods", key = key)
aff <- searchplos("*:*", fl = "title, author_affiliate", key = key)
aff$author_affiliate[[2]]
searchplos("*:*", fl = "id", key = key)
```

here is a list of [options for the search](http://api.plos.org/solr/search-fields/) or can do `data(plosfields); plosfields`

### take a highbrow look!

```{r eval=FALSE}
out <- highplos(q='alcohol', hl.fl = 'abstract', rows=10, , key = key)
highbrow(out)
```

## plots over time
```{r eval = FALSE}
plot_throughtime(terms = "phylogeny", limit = 200, key = key)
```

## is it a boy or a girl? `gender` throughout US history

```{r eval = FALSE}
devtools::install_github("lmullen/gender-data-pkg")
devtools::install_github("ropensci/gender")
```

The gender package allows you access to American data on the gender of names. Because names change gender over the years, the probability of a name belonging to a man or a woman also depends on the *year*:

```{r eval = FALSE}
library(gender)
gender("Kelsey")
gender("Kelsey", years = 1940)
```


