---
title: "cmd101-lecture-ntoes"
author: "gvdr"
date: "October 24, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R for text

Today, after a bit of class management, we are going to cover some fundamentals for interacting with strings. Let's provide ourselves with a variety of character strings.

```{r}
my_string <- "This is a nice little string! Isn't it?"
my_string2 <- 'This little string contains "apices". What is the word for them?'
my_string3 <- "Also this string contains a ' character."
```

Notice the role of the different apices: if we open (and thus close) the string with " we can contain ' as simple text; if we open (and thus close) the string with \' we can contain " as simple text.

Other things that look like an apices are actually not, and they are a tad different:

```{r}
my_string4 <- `my_string`
my_string5 <- 'my_string'

my_string4
my_string5
```

Finally, we can have vectors of strings.

```{r}
library(singer)
places <- unique(singer_locations$city)
```


### The tools of the trade

There are a bunch of libraries which makes life much easier. Two are `stringr` and `glue`, that live in the tidyverse world, another one is `stringi`. `stringi` is the core of `stringr`, and it's **very** powerful and fast. I don't mingle much with text, so I don't use it often: `stringr` does everything I need.

```{r}
library(tidyverse)
library(stringr)
library(glue)
```

#### Gluing is caring

The function `glue()` allows to compose strings in a programmatic and elegant way. I find it more convenient than `paste()` or `paste0()`, but requires to learn a tad of syntax.

`glue()` knows the environment in which you call it. That means that can use variables that were previously defined.

```{r}
My_name <- "Giulio"
My_age <- 32
My_twitter <- "ipnosimmia"
glue('My name is {My_name},',
  ' my age next year is {My_age + 1},',
  ' my twitter handle is @{My_twitter}.')
```

It als preserves indentation and new lines (but you have to start from the beginning):

```{r}
glue('
My name is {My_name},
           my age last year was {My_age - 2},
   my twitter handle is @{My_twitter}.
')
```


If we don't want to create permament variables for those string, we can pass them *after* the string as on-the-fly parameters
_
```{r}
glue('Today is {_____} and the sun shines.',
     today_date = Sys.Date())
```

That's not so nice, is it? We surely could use a `format()`ting. Acciderboli, it would be nice if we could reuse that phrase without having to repast it every time. Well, we can!

```{r}
phrase <- 'Today is {today_date} and the sun shines.'
glue(_____,
     today_date = format(Sys.Date(), "%A, %B %d, %Y"))
```

`glue` also works with data frames, via the `glue_data()` function.

```{r}
singer_locations %>%
 filter(!is.na(city)) %>%
 glue_data("{_____} is from the city of {_____}.") %>%
  head()
```


#### Stringr for the string practitioner

The package `stringr` provides a wide variety of string manipulation functions.

```{r}
getNamespaceExports("stringr")
```


There's a bunch of them starting with `str_`, and they constitute the core of `stringr` functionalities. Let's take a look a `str_view()`:

```{r}
getNamespaceExports("stringr") %>%
  str_view("_____")
```

What happened here? Try to change the string you passed as an input and see what happens.

A similar function is `str_subset()`.

```{r}
getNamespaceExports("stringr") %>%
  str_subset("_____")
```

##### Challenge

Let's use `str_subset()` to identify all the bands' places containing the string `new`.

```{r}
places %>%
  str_subset("____")
```
